---
title: "Sentinel Site Selection Proposal for Genomic Surveillance in Ethiopia (2020-2024)"
author: "Henok T"
# format: html
# For RevealJS, uncomment below and comment out 'format: html'
format: 
  html:
    self-contained: true 
theme: sky
editor: visual
---

```{r include=FALSE}
pacman::p_load(tidyverse, sf, tmap, DT, here)

# Load pre-processed data
dat_multi <- readRDS(here("output/dat_multi.rds"))
dat_multi_sf <- readRDS(here("output/dat_multi_sf.rds"))
eth_adm1 <- readRDS(here("output/eth_adm1.rds"))

knitr::opts_chunk$set(echo = FALSE)
```

## Introduction

-   **Objective**: Propose sentinel sites for genomic surveillance in Ethiopia to enhance outbreak detection and response.
-   **Data Source**: Disease surveillance data (2020-2024) covering 4,852 records from total reporting woredas in the above years.

------------------------------------------------------------------------

-   **Diseases**:
    -   Malaria (`mal`), Meningitis (`mng`), AFP Polio (`afp`), Anthrax (`anth`), Rabies (`rab`), Measles (`msls`): 2020-2024.
    -   Cholera (`chlr`): 2022-2024.
    -   Dengue Fever (`deng_fevr`): 2023-2024.

------------------------------------------------------------------------

-   **Goals**:
    -   Identify high-burden woredas for each disease.
    -   Select woredas with multiple diseases for comprehensive surveillance.
    -   Visualize results interactively for stakeholder use.

## Methodology: Data Collection and Preparation

-   **Scope**: Aggregated disease cases by woreda and region from 2020 to 2024.
-   **Data Limitations**:
    -   Cholera data available only from 2022-2024.
    -   Dengue fever data available only from 2023-2024.
    -   Missing values (`NA`) treated as zero to ensure consistent analysis.

------------------------------------------------------------------------

-   **Processing**:
    -   Summed cases per disease per woreda, respecting data availability periods.
    -   Used `tidyverse` for data cleaning and aggregation.

## Methodology: Site Selection Process

-   **Step 1: Ranking Woredas per Disease**:
    -   For each disease, ranked woredas by case count within each region.
    -   Selected top 20 woredas for Oromia and Amhara (high-population regions).
    -   Selected top 10 woredas for other regions.

------------------------------------------------------------------------

-   **Step 2: Identifying Multi-Disease Woredas**:
    -   Identified woredas appearing in multiple disease rankings.
    -   Calculated:
        -   `num_diseases`: Number of diseases ranked in top 10/20.
        -   `diseases_present`: List of diseases.
        -   `total_cases`: Sum of cases across all diseases (for context).
    -   Filtered for woredas with at least two diseases.
    -   Sorted by region, `num_diseases` (descending), and `total_cases` (descending).

## Methodology: Visualization

-   **Tools**: Used `tmap` for interactive mapping and `DT` for tables(Table isnot working now but will be corrected)
-   **Map Components**:
    -   Woredas colored by `num_diseases` (Blues palette).
    -   Admin 1 (region) boundaries with labels to show geographic context.
    -   Popups with details: region, zone, woreda, number of diseases, diseases present, total cases.

------------------------------------------------------------------------

-   **Table**:Interactive display of proposed woredas with filtering and sorting capabilities.
-   **Shapefiles**:
    -   Admin 3 (woredas): `eth_admbnda_adm3_csa_bofedb_2021.shp`.
    -   Admin 1 (regions): `eth_admbnda_adm1_csa_bofedb_2021.shp`.

# Results: Proposed Sentinel Sites - Interactive Table

-   **Description**:
    -   Shows woredas with multiple diseases, proposed as sentinel sites.
    -   Columns: `region`, `zone`, `woreda`, `num_diseases`, `diseases_present`, `total_cases`.

------------------------------------------------------------------------

-   **How to Access and Use**:
    -   **Filter**: Use the search bar to find specific regions or woredas.
    -   **Sort**: Click column headers (e.g., `num_diseases`) to rank by disease count.
    -   **Navigate**: Use pagination to view all entries.
    -   Helps stakeholders identify priority woredas for surveillance.

## Propose Sentinel Sites - Table {.table-slide .flex-slide}

**Table is not working, needs maintenance by the creator**

```{r}

# DT::datatable(
#   dat_multi %>% 
#     select(region, zone, woreda, num_diseases, diseases_present, total_cases) %>%
#     arrange(region, desc(num_diseases), desc(total_cases)),
#   options = list(
#     pageLength = 5,
#     scrollX = TRUE,
#     searchHighlight = TRUE,
#     # set a fixed height so it fits
#     height = "400px",
#     width = "100%"
#   ),
#   caption = "Table 1: Proposed Sentinel Sites with Multiple Diseases"
# )


```

# Results: Proposed Sentinel Sites - Interactive Table

-   **Description**:

    -   Visualizes proposed sentinel sites (woredas with \>=2 diseases).
    -   Colored by number of diseasese ( darker blue = more disease reported)
    -   Includes Admin 1 boundaries and region names for context.

------------------------------------------------------------------------

-   **How to use the map**:

    -   **Zoom/Pan**: Explore specific regions or woredas.
    -   **Click**:View popups with details (region, zone, woreda, diseases, cases).
    -   **Legend**: Interpret disease count via the color scale.
    -   Ideal for spatial analysis and Site prioritization.

    ```{r map, echo=FALSE}
    #| classes: map-slide 

    tmap::tmap_mode("view")
    tm_shape(dat_multi_sf) +
      tm_polygons(
        "num_diseases",
        title = "Number of Diseases",
        palette = "Blues",
        popup.vars = c("region", "zone", "woreda", "num_diseases", "diseases_present", "total_cases"),
        id = "woreda",
        alpha = 0.7
      ) +
      tm_shape(eth_adm1) +
      tm_borders(
        col = "black",
        lwd = 2,
        alpha = 0.8
      ) +
      tm_text(
        "adm1_en",
        size = 0.8,
        col = "black",
        fontface = "bold"
      ) +
      tm_layout(
        title = "Sentinel Sites with Multiple Diseases (2020-2024)",
        title.position = c("center", "top"),
        # legend.position = c("left", "bottom")
      )


    ```

## Discussion

-   **Key Findings**:
    -   \[Placeholder: e.g., "Oromia and Amhara have multiple woredas with 3+ diseases, ideal for surveillance hubs."\]
    -   Limited cholera and dengue data may underrepresent some woredas' burden.

------------------------------------------------------------------------

-   **Implications**:
    -   Multi-disease woredas ensure broad pathogen coverage.
-   **Limitations**:
    -   Cholera (2022-2024) and dengue (2023-2024) data gaps.
    -   Potential under reporting or missing data (`NA` values).
    -   utilization of cases definitions in healht facilites - hence reported cases might not really be cases.

## Conclusion

-   **Summary**:
    -   Proposed sentinel sites based on high disease burden and diversity.
    -   Interactive table and map generated to facilitate data-driven decisions.
-   **Next Steps**:
    -   Refine the proposed sites by improving selection approach.
    -   Validate sites through discussion with regions and field assessments.
-   **Contact**: Henok T (\[henoktadesse95\@gmail.coml\])
